---
title: "Descriptive analysis"
author: "Silvia Téliz"
date: "2025-05-13"
output: html_document
---

This script matches the skin tone data with the candidate results data. The skin tone data is obtained from the `skintone` dataset, which contains the skin tone labels for candidates' portraits. The candidate results data is obtained from the `dep_cand_results`, `sen_cand_results`, and `pres_cand_results` datasets.

# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages}
library(tidyverse)
library(magrittr)
library(haven)
```

# Import data

```{r data}
# Candidate-results data
dep_cand_results <- read_csv("../data/processed/03_dep_cand_results.csv")
pres_cand_results <- read_csv("../data/processed/03_pres_cand_results.csv")
sen_cand_results <- read_csv("../data/processed/03_sen_cand_results.csv")

# Skin tone measurement data
skintone <- read_csv("../data/processed/04_skintone.csv")

# PRODER data
proder <- read_dta("../data/raw/baseproder2019.dta")

```

# Skin tone data

```{r clean_skintone}
# Reverse tone_label (1 is 11, 11 is 1)
skintone <- skintone %>%
  mutate(tone_label = 12-tone_label)

# Select relevant variables from skintone data. In some images, two faces were detected. We select the one with the highest confidence score.

skintone_ine <- skintone %>%
  filter(source == "ine" & face_id == 1) %>%
  select(file, tone_label, skin_tone)
names(skintone_ine) <- c("id", "skin_ine", "hex_ine")


skintone_sv <- skintone %>%
  filter(source == "sv" & face_id == 1) %>%
  select(file, tone_label, skin_tone)
names(skintone_sv) <- c("id", "skin_sv", "hex_sv")

# Remove .jpg extension from id column
skintone_ine <- skintone_ine %>%
  mutate(id = str_remove(id, ".jpg"))
skintone_sv <- skintone_sv %>%
  mutate(id = str_remove(id, ".jpg"))

```


```{r match}
# Match skin tone data with candidate results

# Deputies dataset
dep_cand_results <- dep_cand_results %>%
  left_join(skintone_ine, by = "id") %>%
  left_join(skintone_sv, by = "id") %>%
  mutate(contest = "dep")

# Senators dataset
sen_cand_results <- sen_cand_results %>%
  left_join(skintone_ine, by = "id") %>%
  left_join(skintone_sv, by = "id") %>%
  mutate(contest = "sen")

# President dataset
pres_cand_results <- pres_cand_results %>%
  left_join(skintone_ine, by = "id") %>%
  left_join(skintone_sv, by = "id") %>%
  mutate(contest = "pres")

```

# PRODER data
```{r proder}
# RGB to HEX dataset
palette <- data.frame(
  hex = c("#654d3e", "#775741", "#876249", "#946c51", "#a0765a", "#a87f64", 
          "#b1886c", "#b69279", "#be9d86", "#c5a691", "#c8ac99"),
  rgb = c("101, 77, 62", "119, 87, 65", "135, 98, 73", "148, 108, 81", "160, 118, 90", "168, 127, 100", 
          "177, 136, 108", "182, 146, 121", "190, 157, 134", "197, 166, 145", "200, 172, 153"),
  label = c(11, 10, 9, 8, 7, 6, 5, 4, 3, 2, 1))

# Separate rgb values into different columns in palette dataset
palette %<>%
  separate(rgb, into = c("r", "g", "b"), sep = ", ") %>%
  mutate(r = as.numeric(r), g = as.numeric(g), b = as.numeric(b))

# Select relevant variables
proder %<>%
  select(ent, p160101_codigohex, r_rgbd, g_rgbd, b_rgbd)

# Compute euclidean distance between RGB values and palette
compute_distance <- function(row_rgb, ref_rgb) {
  sqrt(sum((row_rgb - ref_rgb)^2))
}

# For each observation, find the closest category
proder %<>%
  # Remove rows with NA in any column
  filter(!is.na(r_rgbd) | !is.na(g_rgbd) | !is.na(b_rgbd)) %>%
  rowwise() %>%
  mutate(
    proder_label = {
      obs_rgb <- c(r_rgbd, g_rgbd, b_rgbd)
      distances <- apply(palette[, 2:4], 1, function(ref_rgb) compute_distance(obs_rgb, ref_rgb))
      palette$label[which.min(distances)]
    }
  ) %>%
  ungroup()

# Obtain average skin tone per state
proder_state <- proder %>%
  group_by(ent) %>%
  summarise(skin_state_pop = mean(proder_label, na.rm = TRUE))


#mean(proder$proder_label, na.rm = TRUE) # Mean skin tone is 7.08

```


# Join datasets

```{r join}
# Join all datasets
cand_skin_results <- bind_rows(dep_cand_results, sen_cand_results, pres_cand_results)

# Bring contest column after name column
cand_skin_results %<>%
  select(id, name, contest, everything())

```


```{r new_vars}

# Obtain mean skin tone from both skin tone measures. If one measure is absent, use only existing measure. If both measures are absent, use NA.
cand_skin_results %<>%
  mutate(skin_pp = ifelse(is.na(skin_ine) & is.na(skin_sv), NA,
                            ifelse(is.na(skin_ine), skin_sv,
                                   ifelse(is.na(skin_sv), skin_ine,
                                          (skin_ine + skin_sv) / 2))))

# Create new skin tone variables
# Add mean skin tone per state, match by ent and state_id
cand_skin_results %<>%
  left_join(proder_state, by = c("state_id" = "ent"))


# Mean candidate skin tone per state
cand_skin_results %<>%
  group_by(state_id) %>%
  mutate(skin_state_can = mean(skin_pp, na.rm = TRUE)) %>%
  ungroup()

# Mean candidate skin tone per contest
cand_skin_results %<>%
  group_by(contest) %>%
  mutate(skin_contest_can = mean(skin_pp, na.rm = TRUE)) %>%
  ungroup()

# Mean candidate skin tone per contest and state
cand_skin_results %<>%
  group_by(state_id, contest) %>%
  mutate(skin_contest_state_can = mean(skin_pp, na.rm = TRUE)) %>%
  ungroup()

# Mean candidate skin tone per party
cand_skin_results %<>%
  group_by(party) %>%
  mutate(skin_party_can = mean(skin_pp, na.rm = TRUE)) %>%
  ungroup()


```


# Export

```{r export}
# Export the dataset
write_csv(cand_skin_results, "../data/processed/05_cand_skin_results.csv")

write_csv(palette, "../data/processed/05_palette.csv")

write_csv(proder, "../data/processed/05_proder.csv")

```



