---
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "../products") })

title: "Descriptive analysis"
author: "Silvia Téliz"
date: "2025-05-13"
output:
  html_document:
    toc: true
    toc_depth: 2      
    toc_float: true
---

This script produces the descriptive analysis of the main dataset (`data/processed/cand_skin_results.Rmd`).


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,  include = TRUE, warning = FALSE, message = FALSE)
```


```{r packages}
library(tidyverse)
library(magrittr)
library(readr)
library(ggplot2)
library(waffle)
library(sf)
library(kableExtra)
library(gt)
library(gtExtras)
```


```{r data}
data <- read_csv("../data/processed/05_cand_skin_results.csv")
proder <- read_csv("../data/processed/05_proder.csv")
```



# Image data availability

```{r source_comparison}
# Comparison of missing data by source
missing_all <- data %>%
  group_by(contest, ine_d, sv_d) %>%
  summarize(n = n()) %>%
  mutate(miss_class = case_when(
    ine_d == 0 & sv_d == 0 ~ "Missing in both sources",
    ine_d == 1 & sv_d == 1 ~ "Existing in both sources",
    ine_d == 0 ~ "Missing in INE",
    sv_d == 0 ~ "Missing in SaberVotar")) %>%
  group_by(contest) %>%
  mutate(total_contest = sum(n),
         total = 1242)

# Reorder contest factor levels
missing_all$contest <- factor(missing_all$contest, 
                               levels = c("pres", "sen", "dep"),
                               labels = c("Presidency", "Senate", "Congress"))

# Total number of candidates with missing data in both sources
n_missing <- missing_all %>%
  group_by(miss_class) %>%
  summarize(n = sum(n),
            prop = round((n / 1242)*100, 2))
 
```

**Main takeaways:**

- A total of 21 candidates (1.69%) are missing data from both, their INE profile and SaberVotar profile. These candidates are from the MC party and the PAN-PRI-PRD coalition. They are part of the congress race and senate race. They belong to the states: Mexico, Jalisco, Guanajuato, San Luis Potosi, Coahuila, Oaxaca, Mexico City, Chiapas, Quintana Roo, and Guerrero.
- A total of 144 candidates (11.59%) are missing data from their INE profile.
- A total of 32 candidates (2.58%) are missing data from their SaberVotar profile.
- The SaberVotar data has a lower proportion of missing data. Therefore, a possible solution is to use this data source and impute skin tone measurements from the INE profile picture if needed. Averaging both measures would not be optimal because skin tone difference from both sources is often large.

**Correlations with outcome variable:**


- The existence of an INE profile picture is weakly and positively correlated with electoral success (r=`r round(cor(data$ine_d, data$win, use = "pairwise.complete.obs"), 2)`). 
- The existence of a SaberVotar profile picture is weakly and positively correlated with electoral success (r=`r round(cor(data$sv_d, data$win, use = "pairwise.complete.obs"), 2)`). 
- The existence of a social media profile (a rough proxy for presence in public media environment) is weakly and positively correlated with electoral success (r=`r round(cor(data$social_media_d, data$win, use = "pairwise.complete.obs"), 2)`). 



```{r plot_missing_all}
# Plot missing data by source
ggplot(missing_all, aes(x = contest, y = n, fill = miss_class)) +
  geom_bar(stat = "identity", position = "fill") +
  labs(title = "Image data availability by source and contest",
       x = "Contest",
       y = "Proportion of candidates") +
  theme_classic() +
  theme(plot.margin = margin(10, 25, 10, 10)) +
  theme(panel.grid.major.x = element_line(color = "gray80", size = 0.5)) +
  scale_fill_tableau(palette = "Colorblind 10")



```


# Missing image data {.tabset}
## By state
The following figures show the distribution of missing image data by state. The state with the highest proportion of candidates with missing visual data is Guerrero, followed by Quintana Roo, Chiapas, and Mexico City. 

```{r missing_state, echo = F}
# Missing image data for all contests, by state
missing_state <- data %>%
  group_by(state) %>%
  mutate(na_both = ifelse(ine_d == 0 & sv_d == 0, 1, 0)) %>%
  summarize(n_ine = sum(ine_d == 0),
            prop_ine = round(n_ine / n(), 2),
            n_sv = sum(sv_d == 0),
            prop_sv = round(n_sv / n(), 2),
            n_both = sum(na_both ==1),
            prop_both = round(n_both / n(), 2)) %>%
  arrange(desc(prop_both))

# Capitalize first letter of "state"
missing_state$state <- str_to_title(missing_state$state)

# Print missing_state with kable
missing_state %>%
  arrange(desc(prop_both)) %>%
  select(state, n_both, prop_both) %>%
  # change column names
  rename("State" = state,
         "Number" = n_both,
         "Proportion" = prop_both) %>%
  kable() %>%
  kable_styling("striped", full_width = T) %>%
  add_header_above(c("Candidates with missing image data per state" = 3)) %>% 
  scroll_box(height = "300px")

```


```{r plot_state, echo = F, include = F}
# Plot missing image data for all contests
missing_state %>%
  ggplot(aes(x = reorder(state, -prop_both), y = prop_both)) +
  geom_col() +
  geom_text(aes(label = ifelse(n_both == 0, NA, paste0("n = ", n_both))),
          hjust = -0.5, size = 2.5) +
  coord_flip() +
  labs(title = "Distribution of missing image data by state",
       x = "State",
       y = "Proportion of missing image data") +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 1)) +
  theme_classic() +
  theme(plot.margin = margin(10, 25, 10, 10)) +
  theme(panel.grid.major.x = element_line(color = "gray80", size = 0.5))

```

```{r mex_map, include = F}
mex_map <- st_read("mex.geojson")%>%
  # change first column name
  mutate(state = shapeName)

# Change state names
mex_map$state %<>% as_factor() %>%
  fct_recode("Ciudad De Mexico" = "Distrito Federal",
             "Coahuila" = "Coahuila de Zaragoza",
                      "Michoacan" = "Michoacan de Ocampo",
                      "Queretaro" = "Queretaro de Arteaga",
                      "Veracruz" = "Veracruz de Ignacio de la Llave")

map_data <- mex_map %>%
  left_join(missing_state, by = "state") 

```



```{r map_missing}
ggplot(map_data) +
  geom_sf(aes(fill = prop_both), color = "white", size = 0.2) +
  scale_fill_gradient(low = "gray90", high = "gray10", na.value = "red") +
  theme_void() +  # removes axes, ticks, and background
  theme(legend.position = "right",            # or "bottom", or "none"
    panel.grid = element_blank(),
    axis.ticks = element_blank(),
    axis.text = element_blank(),
    axis.title = element_blank(),
    plot.title = element_text(hjust = 0.5)) +
  theme(plot.margin = margin(10, 25, 10, 10)) +
  labs(title = "Proportion of candidates with missing image data (from both sources)",
       fill = "Proportion")

```


## By contest
Most of the missing data is concentrated in the Congress race (deputies). There is no missing data in the presidantial race. The following figure shows the distribution of missing image data by contest.

```{r missing_contest}
# Missing image data for all contests, by state
missing_contest <- data %>%
  group_by(contest) %>%
  mutate(na_both = ifelse(ine_d == 0 & sv_d == 0, 1, 0)) %>%
  summarize(n_ine = sum(ine_d == 0),
            prop_ine = round(n_ine / n(), 2),
            n_sv = sum(sv_d == 0),
            prop_sv = round(n_sv / n(), 2),
            n_both = sum(na_both ==1),
            prop_both = round(n_both / n(), 2)) %>%
  arrange(desc(prop_both))


```


```{r plot_contest, echo = F}
# Plot missing image data for all contests
missing_contest %>%
  ggplot(aes(x = reorder(contest, -prop_both), y = prop_both)) +
  geom_col() +
  geom_text(aes(label = ifelse(n_both == 0, NA, paste0("n = ", n_both))),
          hjust = -0.5, size = 2.5) +
  coord_flip() +
  labs(title = "Distribution of missing image data by contest",
       x = "Contest",
       y = "Proportion of missing image data") +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 1)) +
  theme_classic() +
   theme(plot.margin = margin(10, 25, 10, 10)) +
  theme(panel.grid.major.x = element_line(color = "gray80", size = 0.5))

```



## By party
Most of the missing data is concentrated in the party "MC" and the coalition "Fuerza y Corazon por Mexico" (pan_pri_prd). The following figure shows the distribution of missing image data by party.

```{r missing_party}
# Missing image data for all contests, by state
missing_party <- data %>%
  group_by(party) %>%
  mutate(na_both = ifelse(ine_d == 0 & sv_d == 0, 1, 0)) %>%
  summarize(n_ine = sum(ine_d == 0),
            prop_ine = round(n_ine / n(), 2),
            n_sv = sum(sv_d == 0),
            prop_sv = round(n_sv / n(), 2),
            n_both = sum(na_both ==1),
            prop_both = round(n_both / n(), 2)) %>%
  arrange(desc(prop_both))
```


```{r plot_party, echo = F}
# Plot missing image data for all contests
missing_party %>%
  ggplot(aes(x = reorder(party, -prop_both), y = prop_both)) +
  geom_col() +
  geom_text(aes(label = ifelse(n_both == 0, NA, paste0("n = ", n_both))),
          hjust = -0.5, size = 2.5) +
  coord_flip() +
  labs(title = "Distribution of missing image data by party",
       x = "Party",
       y = "Proportion of missing image data") +
  scale_y_continuous(expand = c(0, 0), limits = c(0, 1)) +
  theme_classic() +
   theme(plot.margin = margin(10, 25, 10, 10)) +
  theme(panel.grid.major.x = element_line(color = "gray80", size = 0.5))

```


# Skin tone distribution {.tabset}

```{r skin_missing}
missing_skin <- data %>%
  filter(ine_d == 1 | sv_d == 1) %>%
  summarize(n_missing = sum(is.na(skin_pp)),
            prop_existing = round(n_missing / n(), 5),
            prop_total = round(n_missing / nrow(data), 5))

```


**Main takeaways:**

- Skin tone measure is the PRODER scale, which ranges from 1 (lightest tone) to 11 (darkest tone).
- The skin tone classification algorithm was not able to measure the skin tone of `r missing_skin$n_missing` candidates. This number represents `r round(missing_skin$prop_existing*100, 2)`% of the candidates with image data and `r round(missing_skin$prop_total*100, 2)`% of the total candidates in the dataset. This issue may be resolved by adjusting the algorithm's facial detection parameters.
- The overall skin tone distribution is slightly right-skewed, indicating that most candidates have a medium-light skin tone. The mean skin tone is `r round(mean(data$skin_pp, na.rm = TRUE), 2)` and the standard deviation is `r round(sd(data$skin_pp, na.rm = TRUE), 2)`. The skin tone distribution is shown in the following figure.
- The national population mean skin tone is `r round(mean(proder$proder_label),2)` with a standard deviation of `r round(sd(proder$proder_label),2)`, according to PRODER data. Therefore, the mean skin tone of candidates is below the national average by 2.18 points.

## Candidates' skin tone

```{r skin_hist_cand}
# Plot  candidate skin tone distribution 
data %>%
  ggplot(aes(x = skin_pp)) +
  geom_histogram(bins = 11, fill = "gray50", color = "white") +
  geom_vline(xintercept = 4.9, linetype = "solid", color = "black", size = 0.8) +
  annotate("text", x = 4.9, 
           y = 200, label = "Mean = 4.9", 
           vjust = -2, hjust = -0.3, size = 3) +
  annotate("text", x = 4.9, y = 200, 
           label = "SD = 1.92", 
           vjust = -.5, hjust = -0.3, size = 3) +
  labs(title = "Candidate skin tone distribution",
       x = "Skin tone",
       y = "Count") +
  scale_x_continuous(expand = c(0, 0), limit = c(1,11), breaks = 1:11) +
  scale_y_continuous(expand = c(0, 0), limit = c(0, 300)) +
  theme_classic() +
  theme(
    plot.margin = margin(10, 25, 10, 10),
    panel.grid.major.x = element_line(color = "gray80", size = 0.5))

```


## Population's skin tone
```{r skin_hist_pop}
# Plot  candidate skin tone distribution 
proder %>%
  ggplot(aes(x = proder_label)) +
  geom_histogram(bins = 11, fill = "gray50", color = "white") +
  geom_vline(xintercept = 7.08, linetype = "solid", color = "black", size = 0.8) +
  annotate("text", x = 7.08, 
           y = 1500, label = "Mean = 7.08", 
           vjust = -2, hjust = -0.3, size = 3) +
  annotate("text", x = 7.08, y = 1500, 
           label = "SD = 2.09", 
           vjust = -.5, hjust = -0.3, size = 3) +
  labs(title = "Population skin tone distribution",
       x = "Skin tone",
       y = "Count") +
  scale_x_continuous(expand = c(0, 0), limit = c(1,11), breaks = 1:11) +
  scale_y_continuous(expand = c(0, 0), limit = c(0, 1800)) +
  theme_classic() +
  theme(
    plot.margin = margin(10, 25, 10, 10),
    panel.grid.major.x = element_line(color = "gray80", size = 0.5))

```

## By state

```{r skin_state}
# Grouped summary + raw data column
skin_state <- data %>%
  filter(state != "nacional") %>%
  group_by(state) %>%
  summarize(
    n = n(),
    mean = round(mean(skin_pp, na.rm = TRUE), 2),
    sd = round(sd(skin_pp, na.rm = TRUE), 2),
    skintone_dist = list(skin_pp)) 

# Capitalize first letter of "state"
skin_state$state %<>% str_to_title()

skin_state %>%
  gt() %>%
  tab_header(title = "Candidate skin tone distribution by state") %>%
  gtExtras::gt_plt_dist(
    column = skintone_dist,
    type = "boxplot",
    fill = "gray50",
    bw = 0.5)

```

## By contest
```{r skin_contest}
# Grouped summary + raw data column
skin_contest <- data %>%
  group_by(contest) %>%
  summarize(
    n = n(),
    mean = round(mean(skin_pp, na.rm = TRUE), 2),
    sd = round(sd(skin_pp, na.rm = TRUE), 2),
    skintone_dist = list(skin_pp)) %>%
  arrange(desc(mean))

# Recode contest
skin_contest$contest %<>% as_factor() %>%
  fct_recode("Congress" = "dep",
             "Senate" = "sen",
             "Presidency" = "pres")

skin_contest %>%
  gt() %>%
  tab_header(title = "Candidate skin tone distribution by contest") %>%
  gtExtras::gt_plt_dist(
    column = skintone_dist,
    type = "boxplot",
    fill = "gray50",
    bw = 0.5)

```



## By party
```{r skin_party}
# Grouped summary + raw data column
skin_party <- data %>%
  filter(party != "ind") %>%
  group_by(party) %>%
  summarize(
    n = n(),
    mean = round(mean(skin_pp, na.rm = TRUE), 2),
    sd = round(sd(skin_pp, na.rm = TRUE), 2),
    skintone_dist = list(skin_pp)) %>%
  arrange(desc(mean))


skin_party %>%
  gt() %>%
  tab_header(title = "Candidate skin tone distribution by party") %>%
  gtExtras::gt_plt_dist(
    column = skintone_dist,
    type = "boxplot",
    fill = "gray50",
    bw = 0.5)

```

















